---
title: "Calculating Dirichlet Mixtures with Actual Data"
author: "Rob Schick, PhD"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating Dirichlet Mixtures with Actual Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This builds on `dirichletSampler.Rmd`, which was starting to get a bit unwieldy. Here I'm going to simulate a posterior using actual priors and data. For data, I'll start with the data = 0 test that Michail suggests. Then we'll use actual data. Right now I have this function:

```{r calcC}
library(eliciteg)
calcC <- function(data, prior){

  dd <- as.vector(data)
  pp <- as.vector(prior)
  if(length(dd) != length(pp)) stop('Data and prior lengths do not match')
   
  logc <- lgamma(sum(pp)) - sum(lgamma(pp)) - (lgamma(sum(pp + data)) - sum(lgamma(pp + data)))

  return(exp(logc))
}
```

And I have a list (`allqs`) with all the individual priors. We'll make some data with 1s, and show an example of how to use this with two different experts. 

```{r}
dat <- matrix(1, nrow = 9, ncol = 9)
calcC(dat, allqs[[1]][[1]])
calcC(dat, allqs[[1]][[6]])
```

Ok, well, we've learned one thing that with data = 0, than we just get 1 returned from the function, which means that we'll not choose any of the individuals with any probability. Is that right?

Anyway, to get all the c values, I have to run `lapply()` to get the values for each expert-prior / data combination.

```{r}
cvec <- rapply(allqs, function(x) calcC(dat, x), how = 'list')
```

Ok, that's the c values, now we need to assemble K. For example, here's the algorithm that would work in commented code:

```{r, eval=FALSE}
# Get the individual numerators for each of the 8 experts
numexp <- length(cvec[[1]])
up1 <- (1 / numexp) * unlist(cvec[[1]][1])
up2 <- (1 / numexp) * unlist(cvec[[1]][2])
up3 <- (1 / numexp) * unlist(cvec[[1]][3])
up4 <- (1 / numexp) * unlist(cvec[[1]][4])
up5 <- (1 / numexp) * unlist(cvec[[1]][5])
up6 <- (1 / numexp) * unlist(cvec[[1]][6])
up7 <- (1 / numexp) * unlist(cvec[[1]][7])
up8 <- (1 / numexp) * unlist(cvec[[1]][8])

# Get the Standard denominator, which is the same for all 8 experts
down <- (1 / numexp) * sum(unlist(cvec[[1]]))

# Calculate k_i for each individual expert
k1 <- up1 / down
k2 <- up2 / down
k3 <- up3 / down
k4 <- up4 / down
k5 <- up5 / down
k6 <- up6 / down
k7 <- up7 / down
k8 <- up8 / down

# Assemble the K vector that will be used in the multinomial sampling
K <- c(k1, k2, k3, k4, k5, k6, k7, k8)
```

Here's the slightly more R-ish way, in the sense that it is vectorised:

```{r}
numexp <- length(cvec[[1]])
upq1 <- (1 / numexp) * unlist(cvec[[1]])
down <- (1 / numexp) * sum(unlist(cvec[[1]]))
kq1 <- upq1 / down
```

Now I should be able to capture this with `lapply()`, i.e. do it in a really R-ish way. First we right the function. What this function will do is loop over each list element, which is itself a list. Take that list, calculate and then return the values for K as a list. `klist` won't be recrusive; instead it's a 6 element list (one for each question). Each element in the list is a length 8 vector that contains the 8 $k_i$'s for each expert. We'll use each list element when sampling the prior.

```{r calcK}
calcK <- function(cdat){
  numexp <- length(unlist(cdat))
  up <- (1 / numexp) * unlist(cdat)
  down <- (1 / numexp) * sum(unlist(cdat))
  up / down
}

klist <- lapply(cvec, function(x) calcK(x))
```

Next we sample $k_i$, and then to use that index `idx` to choose the prior, which is a list element out of `allqs`. Here we apply this one time for one question using Jim's algorithm.

```{r simpleGibbsfx}
nreg <- length(regID)

idx <- which(rmultinom(1, 1, klist[[1]]) == 1)

prior <- matrix(data = unlist(allqs[[1]][idx]), nrow = length(regID), ncol = length(regID))
di <- matrix(rgamma(nreg * nreg, 
                    shape = dat + prior, 
                    scale = 1), nreg, nreg) 

post <- di / matrix(colSums(di), nreg, nreg, byrow = T)
colnames(post) <- row.names(post) <- regID
round(post, 3)

```

That looks good, and now we need to sample within a Gibbs framework. What I'll do here, and elsewhere is to:

1. Initialise an empty matrix `sumh`
2. set up the number of Gibbs steps `ng`
3. Within the loop, choose an index
4. Use that index to choose a prior
5. Sample from the Dirichlet using Jim's algorithm
5. Add these values to `sumh`
6. After the loop, summarise the posterior by dividing by `ng`

```{r simpleGibbs}
sumh <- post * 0
colnames(sumh) <- row.names(sumh) <- regID
question <- 6
ng <- 10
for(i in 1:ng){
  idx <- which(rmultinom(1, 1, klist[[question]]) == 1)
  # print(idx)
  prior <- matrix(data = unlist(priorList[['females']][idx]), nrow = length(regID), ncol = length(regID))
  di <- matrix(rgamma(nreg * nreg, 
                    shape = dat + prior, 
                    scale = 1), nreg, nreg) 

  post <- di / matrix(colSums(di), nreg, nreg, byrow = T)  
  sumh <- sumh + post
}

movePostprob <- sumh / ng
round(movePostprob, 3)
```

Right. Now I need to do it for all questions. In the Gibbs loop, we loop over months and regions.
